Effects of pulsed hollow electron lens operation on the beam core in HL-LHC: \\First experimental studies and simulations.

### 2017-12-26 resex.tex
update 26 Dec 2017

\begin{figure}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_7thhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_10thhv_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn (left) and vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (right) from simulation results (Distribution Tracking) based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The excitation was applied in both planes with an amplitude of 96~nrad.}
\end{figure}
For the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing the emittance also appears to start at an increased initial value and then stays almost constant. This behavior is due to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium distribution, and as the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}. Besides an increased emittance, the new distribution also becomes non-Gaussian as can be seen from the histograms shown in Fig.~\ref{fig:patternhist} for the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing. This initial fast change of the beam distribution was also to the knowledge of the authors for the first time experimentally observed (Sec.~\ref{sec:simex10} and Sec.~\ref{sec:simex7}), however on a time scale of minutes instead of seconds.

\subsubsection{10th turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing was tested in 2016 with an excitation in the vertical plane only. Experimentally, the following observations were made (Fig.~\ref{fig:10thexp}):
\begin{itemize}
	\item increase of the loss rate with the excitation amplitude,
	\item increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane.
\begin{figure}[t]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.49\linewidth]{bunch_intensity_v10th_no_damper_avg_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emith_avg_rel_v10th_no_damper_annotate.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emitv_avg_rel_v10th_no_damper_annotate.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and emittance (bottom) obtained during the 2016 experiments. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad, the excitation amplitude was linearly increased for each group of 4~bunches (see Sec.~\ref{sec:adt}). The different colored solid lines labeled with $n\Delta A$ indicate the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation as their envelope.}
\end{figure}
Comparing these experimental results with the simulations shown in Fig.~\ref{fig:10thsim} with only $10^{\mathrm{th}}$~turn pulsing (solid line), the simulations predict qualitatively correctly an excitation amplitude dependent loss rate, but fail to predict the excitation amplitude dependent, continuous emittance growth seen experimentally. Instead, the simulations yield a constant emittance after an initial increase during the first $10^4$~turns or first second. So obviously there is an essential ingredient missing. As it turns out, the constant emittance can be changed to an excitation amplitude dependent emittance growth by adding a random dipole noise component with uniform distribution (dotted line in Fig.~\ref{fig:10thsim}). As a first estimate the value obtained in \cite{md1433_noise_top_energy,md_noise_bbLHC} for 6.5~TeV was used and scaled by the magnetic rigidity to the injection energy of 450~GeV, as no noise estimate is available at injection, yielding a maximum kick amplitude at the transverse damper of approximately
\begin{equation}
\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad},
\end{equation}
Adding this random noise component, the simulations then yield correctly an excitation amplitude dependent emittance growth in the vertical plane only after a first fast adjustment of the beam distribution visible as an increased initial value. In addition, also the loss rates increase. As the loss rates are in general underestimated in simulations compared to the experimental results, the discrepancy between simulations and experiment is therefore further reduced.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_emit2_rel.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:10thsim} Relative intensity (top left) and vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$. The solid line indicates an excitation with only $10^{\mathrm{th}}$~turn pulsing in V and the dotted line with $10^{\mathrm{th}}$~turn pulsing in V plus a random dipole noise component in H+V of 6~nrad.}
\end{figure}

@@

\subsubsection{7th turn pulsing\label{sec:simex7}}
The $7^{\mathrm{th}}$~turn pulsing pattern is the only pulsing pattern which has been tested during both experiments in 2016 and 2017 and was intended to serve as a proof for the reproducibility of the results. However, in 2017 the machine tune was changed from (64.28,59.31) to (62.27,60.295) in standard operation. The attempt to change back to the fractional tune of (.28,.31) as chosen in 2016 however lead to large losses during the experiments in 2017 and could thus not be changed. ???This change in tune also entailed a change in the driving resonances (Fig.~\ref{fig:7th2017fma}). As will be detailed in this section, the qualitative evolution of losses and emittance growth are still similar despite the different tunes, but the quantitative results differ.???
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_ord14_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skhv_dp0_ord7.png}
	\end{minipage}	
	\caption{\label{fig:7th2017fma} FMA analysis for $7^{\mathrm{th}}$~turn pulsing based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $7Q_x$ and $7Q_y$ resonance are both excited and in addition the 14th order $7Q_x+7Q_y$.}
\end{figure}

\begin{itemize}
	\item Don't understand why there is emittance growth in H for 2016. There is no emittance growth in the experiment.
	\item repeat simulations for 2017 for all three planes and with random noise => check if there is only emittance growth in V.
	\item experimentally: 2016 + 2017 losses in all cases, only emittance growth in V in 2017
\end{itemize}
\subsubsection{Effect of the transverse damper\label{sec:damp}}
Compare random and nth turn -> no damping for nth, only for random.

@@

\section{Appendixes}

>>>>

\begin{figure}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_7thhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_10thhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from simulation results (Distribution Tracking) based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The excitation was applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here $10^4$~turns) minus the initial distribution (here $10^2$~turns), and the ratio as the final distribution divided by the initial distribution. The red line in the residual shows the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure}
For the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing the emittance also appears to start at an increased initial value and then stays almost constant. This behavior is due to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium distribution. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}. Besides an increased emittance, the new distribution also becomes non-Gaussian as can be seen from the residual in respect to the Gaussian distribution in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing (Fig.~\ref{fig:patternhist}). For comparison, also the distribution in both planes for a random excitation is shown This initial fast change of the beam distribution was also to the knowledge of the authors for the first time experimentally observed (Sec.~\ref{sec:simex10} and Sec.~\ref{sec:simex7}). The timescale of this fast adjustment is experimentally however much slower than predicted in simulations - on the timescale of minutes compared to seconds.

\subsubsection{10th turn pulsing\label{sec:simex10}}
The $10^{\mathrm{th}}$~turn pulsing was tested in 2016 with an excitation in the vertical plane only. In summary, the following experimental results were obtained:
\begin{enumerate}
	\item increase of the loss rate with the excitation amplitude (Fig.~\ref{fig:10thexp}, \ref{fig:10thexpsim}),
	\item increase of the emittance growth with the excitation amplitude in the vertical plane but not in the horizontal plane (Fig.~\ref{fig:10thexp}).
	\item change of the beam distribution due to the resonant excitation (Fig.~\ref{fig:10thexp}, \ref{fig:10thexpprof})
\end{enumerate}
As loss rates, emittance growth and distribution stayed unchanged for the 4~reference bunches, the above observations can be truly associated with the effect of the resonant excitation.
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.49\linewidth]{bunch_intensity_v10th_no_damper_avg_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emith_avg_rel_v10th_no_damper_annotate.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{emitv_avg_rel_v10th_no_damper_annotate.png}
	\end{minipage}	
	\caption{\label{fig:10thexp} Relative intensity (top) and emittance (bottom) obtained during the 2016 experiments. For each maximum amplitude $A_{\mathrm{max}}=48$~nrad and 96~nrad, the excitation amplitude was linearly increased for each group of 4~bunches (see Sec.~\ref{sec:adt}). The different colored solid lines labeled with $n\Delta A$ indicate the average over the 4~bunches with the same excitation amplitude together with the $1\sigma$ standard deviation as their envelope.}
\end{figure}

Besides the determination of the scaling of the loss rate and emittance growth with the excitation amplitude as required for the specification of the hollow electron lens, the most novel observation is the direct measurement of the change of the beam distribution, here with the Beam Synchrotron Radiation Telescope (BSRT) \cite{bsrtprofinj}. As an example for the change in distribution, the vertical profile of a reference bunch and one bunch experiencing the maximum excitation are compared in Fig.~\ref{fig:10thexpprof}. The distribution of the excited bunch clearly changes and becomes non-Gaussian, while the reference bunch stays almost unchanged.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_50.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{profile_v10th_slot_1300.png}
	\end{minipage}
	\caption{\label{fig:10thexpprof} Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2016 experiments. The profiles are taken at the end of the $10^{\mathrm{th}}$~turn pulsing in V. The black line in the plot of the residual is the current profile minus the initial profile and is a measure for the overall change of the distribution. The red line is the current profile minus its Gaussian fit, which gives and indication for the deviation of the distribution from a Gaussian distribution. For details on the analysis see \cite{bsrtprofinj}. The distribution of the reference bunch (left), here slot 50, stays unchanged, while the bunch experiencing the maximum excitation (right), here slot 1300, clearly shows a change to a non-Gaussian distribution.}
\end{figure}

Comparing these experimental results with the simulations including only the $10^{\mathrm{th}}$~turn pulsing (solid line, Fig.~\ref{fig:10thsim}), the simulations predict qualitatively correctly an excitation amplitude dependent loss rate, but fail to predict the excitation amplitude dependent, continuous emittance growth seen experimentally.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_emit2_rel.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerrb2uran1_2e-3_10thV_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:10thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$. The solid line indicates an excitation with only $10^{\mathrm{th}}$~turn pulsing in V and the dotted line with $10^{\mathrm{th}}$~turn pulsing in V plus a random dipole noise component in H+V of 6~nrad.}
\end{figure}
Instead, the simulations yield a constant emittance after an initial increase during the first $10^4$~turns or first second. So obviously there is an essential ingredient missing. As it turns out, the constant emittance can be changed to an excitation amplitude dependent emittance growth by adding a random dipole noise component with uniform distribution (dotted line in Fig.~\ref{fig:10thsim}). As a first estimate the value obtained in \cite{md1433_noise_top_energy,md_noise_bbLHC} for 6.5~TeV was used and scaled by the magnetic rigidity to the injection energy of 450~GeV, as no noise estimate is available at injection. This yields a maximum kick amplitude at the transverse damper of approximately
\begin{equation}
\theta_{\mathrm{random,ADT,max}}(\mathrm{450~GeV}) = 6~\mathrm{nrad},
\end{equation}
Adding this random noise component, the simulations then yield correctly an excitation amplitude dependent emittance growth in the vertical plane only after a first fast adjustment of the beam distribution visible as an increased initial value.

Adding a random noise component has in addition the effect of increasing the loss rate. As the loss rates are in general underestimated in simulations compared to the experimental results, the discrepancy between simulations and experiment is therefore further reduced, but a large difference still prevails (Fig.~\ref{fig:10thexpsim}).
\begin{figure}[h]
		\centering
		\includegraphics[width=1.0\linewidth]{scale_amp_10v_ran_10V_sim.png}
	\caption{\label{fig:10thexpsim} Comparison of scaling of loss rates with the $10^{\mathrm{th}}$~turn pulsing excitation amplitude. The experimental results are shown in black (maximum excitation with 48~nrad) and blue (maximum excitation with 96~nrad) and the simulation results (2016 injection optics, standard errors, $(Q_x,Q_y)=(64.28,59.31)$, $10^{\mathrm{th}}$~turn pulsing in V plus random dipole noise in H+V of 6~nrad) in red.}
\end{figure}

@@

\subsubsection{7th turn pulsing\label{sec:simex7}}
The $7^{\mathrm{th}}$~turn pulsing pattern is the only pulsing pattern which has been tested during both experiments in 2016 and 2017 and was intended to serve as a proof for the reproducibility of the results. However, in 2017 the machine tune was changed from (64.28,59.31) to (62.27,60.295) in standard operation accompanied by a slight change in optics. The attempt to change back to the fractional tune of (.28,.31) as chosen in 2016 however lead to large losses during the experiments in 2017. This change in tune entailed a change in the driving resonances, from previously the $7Q_x$ resonances in 2016  (Fig.~\ref{fig:patternfma}) to the $7Q_x$, $7Q_y$ and $7Q_x + 7Q_y$ resonance in 2017 (Fig.~\ref{fig:7th2017fma}) and in addition also the effected amplitudes of the beam distribution.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_ord14_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerrut7skhv_dp0_ord7.png}
	\end{minipage}	
	\caption{\label{fig:7th2017fma} FMA analysis for $7^{\mathrm{th}}$~turn pulsing based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295). The excitation is 96~nrad in both planes. The $7Q_x$ and $7Q_y$ resonance are both excited and in addition the 14th order $7Q_x+7Q_y$.}
\end{figure}
Therefore the effects on the beam also differ for both experiments (see Table~\ref{tab:7thexp} for a general overview and Fig.~\ref{} and Appendix~\ref{} for more details).
\begin{table}[h]
	\caption{\label{tab:7thexp}%
		Overview of the effect of $7^{\mathrm{th}}$~turn pulsing during the two resonant excitation experiments in 2016 and 2017~\cite{resexmd2016,resexmd2017}. The plane of the excitation is abbreviated with H for horizontal, V for vertical and H+V for horizontal and vertical at the same time.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & Experiment 2017 & Experiment 2016  \\
			\colrule
			tune $(Q_x,Q_y)$ & (62.27,60.295) & (64.28,59.31)\\\hline
			excitation plane  & H,V and H+V&  H\\
			losses & similar loss rates for  & losses for pulsing \\
			& pulsing in H,V & in H \\
			&  and H+V & \\
			hor. emit. growth & no & yes, very small \\
			vert. emit. growth & yes, for pulsing in V & no\\
			&  and H+V & \\
		\end{tabular}
	\end{ruledtabular}
\end{table}
Simulations more or less fail completely (Fig.~\ref{fig:7thsim}). Higher amplitudes have been used for simulations to see noticable effect.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_rel_emit2.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016+2017injerra2b2uran1_2e-3_7th_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:7thsim} Relative intensity (top left) and relative vert. emittance (top right), and hor. (bottom left) and vert. (bottom right) emittance obtained from simulations (Distribution Tracking) based on the 2016 injection optics with standard errors and $(Q_x,Q_y)=(64.28,59.31)$ and 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates an excitation with only a random dipole noise component in H+V of 6~nrad, and the dotted and dashed line the results for $7^{\mathrm{th}}$~turn pulsing plus a random dipole noise component in H+V of 6~nrad.}
\end{figure}

\begin{itemize}
	\item Don't understand why there is emittance growth in H for 2016. There is no emittance growth in the experiment.
	\item repeat simulations for 2017 for all three planes and with random noise => check if there is only emittance growth in V \textcolor{red}{done}.
	\item experimentally: 2016 + 2017 losses in all cases, only emittance growth in V in 2017
\end{itemize}
\subsubsection{Effect of the transverse damper\label{sec:damp}}
Compare random and nth turn -> no damping for nth, only for random.

@@

\section{Appendixes}
\section{Results of $7^{\mathrm{th}}$~turn pulsing during 2016 and 2017 experiments}
